---
title: "Child Mortality and Fertility: Two Sides of the Demographic Transition"
author: "Alexander Temerev"
date: "10 June 2026"
output:
  html_document:
    self_contained: true
    theme: cosmo
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 9.5, fig.height = 5.5, dpi = 110
)
library(readr)
library(dplyr)
library(ggplot2)
```

## Introduction

This report explores the relationship between two country-level indicators from
[Gapminder](https://gapminder.org/data) for the year **2020**:

- **Child mortality** — deaths of children under five years of age, per 1,000 live births;
- **Total fertility rate** — average number of children born per woman.

These two indicators are the core of the *demographic transition*: as fewer children die,
families have fewer children. The analysis below maps both indicators across the world and
examines how tightly they are related.

```{r load-data}
mortality <- read_csv(
  "data/ddf--datapoints--child_mortality_0_5_year_olds_dying_per_1000_born--by--country--time.csv"
) |>
  rename(child_mortality = child_mortality_0_5_year_olds_dying_per_1000_born)

fertility <- read_csv(
  "data/ddf--datapoints--children_per_woman_total_fertility--by--country--time.csv"
) |>
  rename(fertility = children_per_woman_total_fertility)

countries <- read_csv("data/countries_lookup.csv") |>
  filter(`is--country`) |>
  select(country, name, region = world_4region)

dat2020 <- mortality |>
  inner_join(fertility, by = c("country", "time")) |>
  filter(time == 2020) |>
  inner_join(countries, by = "country") |>
  mutate(region = recode(region,
    africa = "Africa", americas = "Americas",
    asia = "Asia & Pacific", europe = "Europe & Central Asia"
  ))

# Harmonize Gapminder country names with the map regions used by the maps package
map_name_fixes <- c(
  "United States"                  = "USA",
  "United Kingdom"                 = "UK",
  "UAE"                            = "United Arab Emirates",
  "Congo, Dem. Rep."               = "Democratic Republic of the Congo",
  "Congo, Rep."                    = "Republic of Congo",
  "Cote d'Ivoire"                  = "Ivory Coast",
  "Kyrgyz Republic"                = "Kyrgyzstan",
  "Lao"                            = "Laos",
  "Slovak Republic"                = "Slovakia",
  "St. Lucia"                      = "Saint Lucia",
  "St. Kitts and Nevis"            = "Saint Kitts",
  "St. Vincent and the Grenadines" = "Saint Vincent",
  "Micronesia, Fed. Sts."          = "Micronesia",
  "Eswatini"                       = "Swaziland",
  "Trinidad and Tobago"            = "Trinidad",
  "Antigua and Barbuda"            = "Antigua"
)

dat_map <- dat2020 |>
  mutate(map_region = recode(name, !!!map_name_fixes))

world <- map_data("world") |> filter(region != "Antarctica")

world_indicators <- world |>
  left_join(dat_map, by = c("region" = "map_region"))

theme_worldmap <- theme_void(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2.2, "cm"),
    legend.key.height = unit(0.35, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
  )
```

## Map 1: Child mortality (2020)

```{r map-mortality}
ggplot(world_indicators, aes(long, lat, group = group, fill = child_mortality)) +
  geom_polygon(colour = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "magma", direction = -1, trans = "log10",
    breaks = c(2, 5, 10, 25, 50, 100),
    na.value = "grey85",
    name = "Deaths per 1,000 born (log scale)"
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Child mortality, 2020",
    subtitle = "Deaths of children under age 5 per 1,000 live births | Source: Gapminder"
  ) +
  theme_worldmap
```

In 2020, child mortality ranged from about 2 deaths per 1,000 births in the safest
countries (Iceland, Finland, Singapore, Japan) to more than 100 per 1,000 in parts of
West and Central Africa — a fifty-fold gap. The darkest band on the map runs across
the Sahel and Central Africa, while Europe, East Asia, and North America are uniformly light.

## Map 2: Total fertility rate (2020)

```{r map-fertility}
ggplot(world_indicators, aes(long, lat, group = group, fill = fertility)) +
  geom_polygon(colour = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "viridis", direction = -1,
    breaks = 1:7,
    na.value = "grey85",
    name = "Children per woman"
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Total fertility rate, 2020",
    subtitle = "Average number of children born per woman | Source: Gapminder"
  ) +
  theme_worldmap
```

The fertility map is strikingly similar to the mortality map: the same Sahel and
Central African belt where child mortality is highest is also where women have the
most children (5–7 per woman in Niger, Chad, Somalia, and the DR Congo), while most of
Europe and East Asia are below the replacement level of about 2.1 children per woman.

## Relationship between the two indicators

```{r scatter, fig.height=6}
region_colours <- c(
  "Africa" = "#2EB6E0", "Americas" = "#7CBF4B",
  "Asia & Pacific" = "#E8506E", "Europe & Central Asia" = "#F2C12E"
)

cor_log <- cor(log10(dat2020$child_mortality), dat2020$fertility)

ggplot(dat2020, aes(child_mortality, fertility)) +
  geom_smooth(method = "lm", colour = "grey55", linetype = "dashed", se = FALSE) +
  geom_point(aes(fill = region), shape = 21, colour = "grey25", size = 3, alpha = 0.85) +
  scale_x_log10(breaks = c(2, 5, 10, 25, 50, 100)) +
  scale_fill_manual(values = region_colours, name = NULL) +
  labs(
    title = "Where more children die, more children are born",
    subtitle = sprintf(
      "Each point is a country, 2020 | r = %.2f (fertility vs log child mortality) | Source: Gapminder",
      cor_log
    ),
    x = "Child mortality (deaths per 1,000 born, log scale)",
    y = "Total fertility rate (children per woman)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

### Description and interpretation

The scatterplot shows a strong positive association between child mortality and
fertility: the correlation between the fertility rate and the logarithm of child
mortality is **r = `r sprintf("%.2f", cor_log)`** across `r nrow(dat2020)` countries.
Countries where fewer than 10 children per 1,000 die before age five almost all have
fertility below three children per woman, while nearly every country with mortality
above 50 per 1,000 has fertility above four.

This is the classic signature of the **demographic transition**. Several mechanisms
link the two indicators:

- **Replacement and insurance effects.** Where child death is common, families have
  more children to compensate for expected losses and as insurance for old age.
- **Shared underlying drivers.** Poverty, limited access to healthcare and
  contraception, and lower female education raise both child mortality and fertility
  at the same time, so part of the correlation reflects common causes rather than a
  direct effect of one indicator on the other.
- **Transition dynamics.** Historically, mortality falls first and fertility follows
  with a delay of a generation or two. The countries in the upper-right of the plot
  (mostly in the Sahel) are at an earlier stage of this transition, not on a different
  path — Gapminder's time series show them moving along the same curve that Europe and
  East Asia traversed in the 20th century.

It is worth stressing that **correlation is not causation** here: reducing child
mortality does not mechanically reduce fertility, but no country has achieved low
fertility while child mortality remained high, which is why child survival is widely
seen as a precondition for the fertility transition.

## Personal reflection on the AI workflow

I completed this workshop using an AI coding agent (Claude Code) in the R track. A few
observations about the process:

- **What worked well.** The agent handled the entire pipeline — installing R and pandoc,
  locating the Gapminder indicator files, downloading the CSVs, and writing the R Markdown
  document — with very little intervention from me. Letting the AI choose the indicator
  pair produced a more interesting analysis (mortality vs fertility) than the default
  GDP-vs-life-expectancy example.
- **Where human oversight mattered.** Country names differ between Gapminder and the R
  `maps` package ("United States" vs "USA", "Congo, Dem. Rep." vs "Democratic Republic
  of the Congo"), and an unsupervised join would have silently dropped a dozen countries
  from the maps. Checking the unmatched-country list and adding an explicit recode table
  was the kind of data-quality step that is easy to miss if you accept AI output without
  verifying it.
- **Takeaway.** The AI was strongest as an *executor* — fast, fluent in ggplot2 idioms,
  good at boilerplate — while the judgment calls (which indicators, log scale on
  mortality, the caveat about confounding) are where the human in the loop adds value.
  Iterating with the AI felt less like delegating and more like pair-working with a very
  fast junior analyst.

---

*Data: [Gapminder](https://gapminder.org/data) (free data from World Bank, UN
Population Division, and Gapminder estimates, CC-BY). Analysis: R `r getRversion()`,
ggplot2, knitted with R Markdown.*
